Level-Set method on \texttt{p4est}

### 2015-01-28 paper/tex/abstract.tex
added time evolution figures for stefan problem
added frederic's corrections. Some points still need to be discussed

%!TEX root = draft.tex
\begin{abstract}
Abstract will be filled in after finalization of the paper.
\end{abstract}

\begin{keyword}
Level-set Method \sep Adaptive Grids \sep Parallel Computing \sep Oct-/Quadtree datastructures
\end{keyword}

\maketitle

>>>>

%!TEX root = draft.tex
\begin{abstract}
We present scalable parallel algorithms for the level-set technology on adaptive Quadtree and Octree Cartesian grids. The algorithms are based on a domain decomposition technique and implemented using \texttt{MPI} and the open-source \texttt{p4est} library. An important contribution is a scalable parallel semi-Lagrangian method which, similar to its serial implementation, is free of any time-step restrictions. This is achieved by introducing a scalable global interpolation scheme on adaptive Quadtree and Octree grids. Moreover, we present a simple parallel reinitialization scheme using the pseudo-time transient formulation. Both parallel algorithms scale on the Stampede supercomputer, where we are limited to 4096 cores at most. Finally a relevant application of the algorithms is presented in modeling the crystallization phenomena by solving a Stefan problem, illustrating a level of detailed calculation that would be impossible without a parallel adaptive strategy. We believe that the algorithms presented in this article will be of interest and useful to researchers working with the level-set framework and modeling multi-scale physics in general.
\end{abstract}
\begin{keyword}
Quadtree/Octree Grids \sep Parallel Computing \sep Space Filling Curves \sep Semi-Lagrangian Method \sep Level-set Method
\end{keyword}
\maketitle

### 2015-01-28 paper/tex/applications.tex
added frederic's corrections. Some points still need to be discussed

%!TEX root = draft.tex
\section{Application to Stefan Problem} \label{sec:application}

@@

In this section we apply our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt, and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into \mohammad{How about: Consider a computational domain $\Omega$ decomposed into ...} \comment{this would be the exact phrasing of another article} two subdomains $\Omega^-$ and $\Omega^+$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega^+$ and $T_l$ in the liquid phase $\Omega^-$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega^-, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega^+. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \mathbf{n}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\mathbf{V} \cdot \mathbf{n}),
\end{equation}
where $\kappa$ is the local interface curvature, $V$ is the velocity of the interface, $\mathbf{n}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the curvature and kinetic undercooling coefficients. The interface velocity $V$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\mathbf{V} \cdot \mathbf{n}) = - \left[ D_l \pd{T_l}{\mathbf{n}} - D_s \pd{T_s}{\mathbf{n}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \mathbf{V} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in algorithm \ref{alg:stefan}.

@@

The implementation of the Stefan problem relies on the components described in the previous sections, and it is therefore a good summary of the performance of the various algorithms. We monitored the performance of the code over five time iterations, as presented in algorithm \ref{alg:stefan}, for two different maximum resolutions. In both cases, the forest is built on a $20\times20\times20$ macro-mesh. The maximum tree resolution for the small test is 9, leading to approximately 7M grid points, and the maximum resolution for the large test is 11, corresponding to 105M grid points. The results are presented in figure \ref{fig:stefan_scaling}. As expected from the results obtained for each component in the previous sections, our implementation of the Stefan problem exhibits very satisfactory scaling (c.f.\ table \ref{tab:scaling_stefan}).

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_small.pdf}}
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) octrees with roughly 7M and 105M grid points, respectively. The solid dashed line represents perfect scaling. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential of our algorithms.}
\label{fig:stefan_scaling}
\end{figure}

@@

\subsection{Numerical experiments}
\mohammad{too many transitive statements and keyword ``choose''. Can we rephrase this?}
We now present the results from a large simulation of the Stefan problem on a $20\times20\times20$ macro-mesh and with level-10 Octrees. We choose to define the Gibbs-Tompson anisotropy undercooling coefficients \eqref{eq:stefan_gibbs_tompson} as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\mathbf{n}$ and the x-axis in the $(x,y)$ plane, $\theta_2$ the angle between $\mathbf{n}$ and the x-axis in the $(x,z)$ plane, and
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2).
\end{align*}
Note that these expressions are arbitrary and lead to a variety of shapes. The computation is initialized with twenty spherical seeds of radius $0.0015$ placed randomly in the domain. We choose $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

The simulation was ran on $256$ mpi processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid}, \ref{fig:stefan_crystals} and \ref{fig:stefan_evolution}. The final iteration of the simulation consisted of $167$M grid points when a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points ! Our simulation used only $0.001948\%$ \mohammad{I think if we are going to report percentage, we should trim it to 1st digit} of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the potential of our approach.

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_1.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_2.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_4.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_3.png}}
\caption{Representation of the computational domain (a) and of various crystal observed for the Stefan problem simulation.} \label{fig:stefan_crystals}
\end{center}
\end{figure}

>>>>

%!TEX root = draft.tex
\section{Application to the Stefan Problem} \label{sec:application}

@@

In this section we apply our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt, and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into \mohammad{How about: Consider a computational domain $\Omega$ decomposed into ...} \comment{this would be the exact phrasing of another article} two subdomains $\Omega_l$ and $\Omega_s$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega_s$ and $T_l$ in the liquid phase $\Omega_l$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega_l, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega_s. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \underline{\mathbf{n}}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}:
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\underline{\mathbf{V}} \cdot \underline{\mathbf{n}}),
\end{equation}
where $\kappa$ is the local interface curvature, $\underline{\mbf{V}}$ is the velocity of the interface, $\underline{\mathbf{n}}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the surface tension and kinetic undercooling coefficients. The interface velocity $\underline{\mbf{V}}$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\underline{\mathbf{V}} \cdot \underline{\mathbf{n}}) = - \left[ D_l \pd{T_l}{\underline{\mathbf{n}}} - D_s \pd{T_s}{\underline{\mathbf{n}}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \underline{\mathbf{V}} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in Algorithm \ref{alg:stefan}.

@@

The implementation of the Stefan problem relies on the components described in the previous sections, and it is therefore a good synthesis of the performance of the various algorithms. We monitored the performance of the code over five time iterations, as presented in Algorithm \ref{alg:stefan}, for two different maximum resolutions. In both cases, the forest is built on a $20\times20\times20$ macro-mesh. The maximum tree resolution for the small test is 9, leading to approximately 7M grid points, and the maximum resolution for the large test is 11, corresponding to 105M grid points. The results are presented in figure \ref{fig:stefan_scaling}. As expected from the results obtained for each component in the previous sections, our implementation of the Stefan problem exhibits very satisfactory scaling (c.f.\ table \ref{tab:scaling_stefan}).

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_small.pdf}}
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) Octrees with roughly 7M and 105M grid points, respectively. The solid dashed line represents perfect scaling. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential of our algorithms.}
\label{fig:stefan_scaling}
\end{figure}

@@

\subsection{Numerical experiments}
We now present the results from a large simulation of the Stefan problem on a $20\times20\times20$ macro-mesh and with level-10 Octrees. The Gibbs-Tompson anisotropy undercooling coefficients in equation \eqref{eq:stefan_gibbs_tompson} are defined as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\underline{\mathbf{n}}$ and the x-axis in the $(x,y)$ plane and $\theta_2$ the angle between $\underline{\mathbf{n}}$ and the x-axis in the $(x,z)$ plane. The coefficients
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2),
\end{align*}
are used to enforce a variety of crystal shapes. The computation is initialized with twenty spherical seeds of radius $1.5 \cdot 10^{-3}$ placed randomly in the domain. We take the diffusion coefficients $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

The simulation was ran on $256$ MPI processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid}, \ref{fig:stefan_crystals} and \ref{fig:stefan_evolution}. The final iteration of the simulation consisted of $167$M grid points whereas a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points. Our simulation used only $0.002\%$ of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the ability of our approach to resolve small scale details, while accounting for long range interactions.

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_1.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_2.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_4.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_crystal_3.png}}
\caption{Representation of the computational domain (a) as well as various developped crystal structures obtained from solving the Stefan problem simulation (b,c,d).} \label{fig:stefan_crystals}
\end{center}
\end{figure}

### 2015-01-28 paper/tex/conclusion.tex
added frederic's corrections. Some points still need to be discussed

%!TEX root = draft.tex
\section{Conclusions}
In this article we have presented parallel algorithms for the advection and reinitialization of the level-set functions on adaptive Quadtree and Octree grids using domain decomposition approach. These algorithms are implemented using a combination of \texttt{MPI} and the open-source \texttt{p4est} library. An important feature of the semi-Lagrangian method is its unconditional stability property which must be preserved in parallel, requiring a parallel interpolation scheme. However, due to the non-uniform distribution of departure points, scalable implementation of such algorithm is not trivial on adaptive Quadtree and Octree grids. An asynchronous interpolation algorithm is presented using non-blocking point-to-point communications which proves good scalability.

The scalability of the semi-Lagrangian algorithm, however, depends on the CFL number. Great scalability is observed for intermediate CFL numbers while, e.g. $\text{CFL}\sim10$. At higher CFL numbers, however, the departure points are potentially further dispersed across processors which limits the scalability. This is because the domain decomposition technique used here is based on the Z-ordering of cells and does not take the velocity field information into account. A possible remedy for this problem could be assigning weights to cells based on some estimate of what the grid should look like after one step of the advection algorithm, e.g. by using a forward-in-time integration of grid points. Such an estimate could also reduce the number of semi-Lagrangian iterations. These ideas are postponed for further investigations. We have also presented a simple parallelization technique for the reinitialization algorithm based on the pseudo-time transient formulation. Both the semi-Lagrangian and the reinitialization algorithms show good scalability up to 4096 processors.

Finally, an application of these algorithms is presented in modeling the solidification process by solving a Stefan problem. This application clearly illustrates the applicability of our algorithms to complex multi-scale problems that cannot be treated using the normal domain decomposition techniques on uniform grids. We believe that our findings would be of interest to other researchers who use the level-set framework for similar complex and multi-scale problems.

>>>>

%!TEX root = draft.tex
\section{Conclusions}
In this article we have presented parallel algorithms related to the level-set technology on adaptive Quadtree and Octree grids using a domain decomposition approach. These algorithms are implemented using a combination of \texttt{MPI} and the open-source \texttt{p4est} library. In order to preserve the unconditional stability property of the semi-Lagrangian scheme while enabling scalable computations, we introduced an asynchronous interpolation algorithm using non-blocking point-to-point communications, and demonstrated its scalability.

In particular we showed that the scalability of the semi-Lagrangian algorithm, depends on the CFL number. Great scalability is observed for intermediate CFL numbers, e.g. $\text{CFL}\sim10$. At higher CFL numbers, however, the departure points are potentially further dispersed across processors, which limits the scalability. This is because the domain decomposition technique used here is based on the Z-ordering of cells and does not take the velocity field information into account. A possible remedy for this problem could be assigning weights to cells based on some estimate of the grid structure after one step of the advection algorithm, e.g. by using a forward-in-time integration of grid points. Such an estimate could also reduce the number of semi-Lagrangian iterations. These ideas are postponed for further investigations. We have also presented a simple parallelization technique for the reinitialization algorithm based on the pseudo-time transient formulation. Both the semi-Lagrangian and the reinitialization algorithms show good scalability up to 4096 processors.

Finally, an application of these algorithms is presented in modeling the solidification process by solving a Stefan problem. This application clearly illustrates the applicability of our algorithms to complex multi-scale problems that cannot be treated practically using the domain decomposition techniques on uniform grids. We believe that our findings can serve as a basis to simulating a wide range of multi-scale problems and free body problems.
